The semantic drift of quotations in blogspace: a case study in short-term cultural evolution

### 2016-07-27 brains-protocol.tex
feat/results: more work on results in variation

We say that a word is \emph{substitutable} if it appears in a quote which undergoes a substitution, whether the substitution operates on that word or on another one.
\newtext{For a given group of words $g$, susceptibility is computed as the ratio of the number of times $s_g$ words of that group are substituted to the number of times $s_g^0$ words of that group would be substituted if substitutions fell randomly on substitutable words, that is:}\footnote{
\newtext{$s_g^0$ is computed by summing, over all quotes undergoing a substitution, the ratio of the number of words in a quote that are from group $g$ to the number of words in the same quote that could have been the substitution's target.
If, for instance, half the content words of a given quote are from group $g$, such a quote contributes .5 to the total $s_g^0$.}
}
$$\sigma_g = \frac{s_g}{s_g^0}$$

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/computed-figures/all-susceptibilities-pos.png}
    \caption{\newtext{\textbf{Substitution susceptibility for POS categories:} categories are simplified from the TreeTagger tag set: \emph{J} means adjective, \emph{N}~noun, \emph{R}~adverb, \emph{V}~verb, and \emph{Stopword-like} gathers together all the categories partially or completely affected by the stopword filter (see main text for details).
    The top panel shows the actual $s_{POS}$ and $s_{POS}^0$ counts, the bottom panel shows $\sigma_{POS}$, the ratio of the two.
    Confidence intervals are computed with the Goodman method for multinomial proportions.}
    \todo{Split into 2 subfigures; fix labels; add H0 dashed line.}}
    \label{fig:pos-susceptibilities}
\end{figure}

@@

\begin{figure*}[!th]
    \centering
    \includegraphics[width=\textwidth]{images/computed-figures/paper-susceptibilities-quantilebins_global.png}
    \caption{\textbf{Substitution susceptibility for feature values:} susceptibility to substitution versus feature value of a candidate word for substitution \newtext{(binned by quartiles)}, with 95\% asymptotic confidence intervals \newtext{(Goodman-based multinomial)}.
    }
    \label{fig:feature-susceptibilities-global}
\end{figure*}

@@

Second, the comparison with $\nu_{\phi}^0$ shows that there are two classes of attractors, depending on whether:
\begin{APAenumerate}
\item there is a triple intersection (of $y=x$, $\nu_{\phi}^0$ and $\nu_{\phi}$);
\item or $\nu_{\phi}$ always remains above or below $\nu_{\phi}^0$.
\end{APAenumerate}
The first class (Number of phonemes and Number of synonyms) are features for which the substitution process only brings words slightly closer to $\nu_{\phi}^0$, and no uniform bias can be observed.

On the other hand, the second class (comprising Word frequency, Age of acquisition, and Clustering coefficient) are features for which the substitution process has a clear bias, positive or negative, with respect to the purely random situation ($\mathcal{H}_0$).

Word frequency, with $\nu_{\phi}$ always significantly above $\nu_{\phi}^0$, exhibits a strong bias towards more frequent words. This, in turn, is consistent with the hypothesis that substitution is a recall process, since common words are favored over awkward ones, while it goes against the idea that it could be a familiarity process, where awkward terms would be favored.

Age of acquisition and Clustering coefficient, on the other hand, exhibit a clear negative bias for the substitution process. Both curves are significantly below their respective $\nu_{\phi}^0$ values, which is consistent with the literature on recall: words learned earlier and words with lower clustering coefficient are easier to produce than average~\citep{nelson2013activation,Zevin02}.
Clustering coefficient has the additional particularity that, on average, the destination word does not depend on the start word; that is on average, substitutions will always produce words with a clustering coefficient around $\exp(-2.4) \simeq .1$.

To make things concrete, here is an example substitution taking place in the data set.
At the end of January 2009, many media websites reported the following quote,

\begin{quote}
    ``The massive economic upheaval being experienced across the globe is sparing no one in the consumer electronics world.''
\end{quote}
and a smaller number of media websites, and blogs, reported the following,
\begin{quote}
    ``The massive economic upheaval being experienced across the \textbf{world} is sparing no one in the consumer electronics world.''
\end{quote}
The word \emph{globe} is acquired at an average of 6.5 years old, appears about 3.5k times in the data set, and has a Clustering coefficient of $.24$.
The word it was replaced with, \emph{world}, is acquired on average at 5.3 years old, appears about 146k times in the data set, and has a Clustering coefficient of $.05$. (Both words have four phonemes.)
Such a change, though minor in appearance, is a typical example of alteration along the lines shown by our results.

%globe (age 6.5, phonemes 4.0, freq 8.18, cc -1.42, susceptibility 1.0)
%world  [age 5.32, phonemes 4.0, freq 11.89, cc -2.95, susceptibility 0.272727272727]

@@

\begin{figure*}[!th]
    \centering
    \includegraphics[width=\textwidth]{images/computed-figures/paper-variations-quantilebins_sentencerel.png}
    \caption{}
    \label{fig:feature-variations-sentencerel}
\end{figure*}

>>>>

We say that a word is \emph{substitutable} if it appears in a quote which undergoes a substitution, whether the substitution operates on that word or on another one.
% utiliser "word-level measures", et/ou "For a given value of a given type of word-level measure (say, words of length 5), we define susceptibility as the ratio (...)"
\newtext{For a given group of words $g$ \todo{[not clear this is not a given word -- maybe present by example instead of generally]}, susceptibility is computed as the ratio of the number of times $s_g$ words of that group are substituted to the number of times $s_g^0$ words of that group would be substituted if substitutions fell randomly on substitutable words, that is:}\footnote{
\newtext{$s_g^0$ is computed by summing, over all quotes undergoing a substitution, the ratio of the number of words in a quote that are from group $g$ to the number of words in the same quote that could have been the substitution's target.
If, for instance, half the content words of a given quote are from group $g$, such a quote contributes .5 to the total $s_g^0$.}
}
$$\sigma_g = \frac{s_g}{s_g^0}$$

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/computed-figures/all-susceptibilities-pos.png}
    \caption{\newtext{\textbf{Substitution susceptibility for POS categories:} categories are simplified from the TreeTagger tag set: \emph{J} means adjective, \emph{N}~noun, \emph{R}~adverb, \emph{V}~verb, and \emph{Stopword-like} gathers together all the categories partially or completely affected by the stopword filter (see main text for details).
    The top panel shows the actual $s_{POS}$ and $s_{POS}^0$ counts, the bottom panel shows $\sigma_{POS}$, the ratio of the two.
    Confidence intervals are computed with the Goodman method for multinomial proportions.}
    \todo{Split into 2 subfigures; fix labels; use log-y for second pane; add H0 dashed line.}}
    \label{fig:pos-susceptibilities}
\end{figure}

@@

\begin{figure*}[!th]
    \centering
    \includegraphics[width=\textwidth]{images/computed-figures/paper-susceptibilities-quantilebins_global.png}
    \caption{\textbf{Substitution susceptibility for feature values:} susceptibility to substitution versus feature value of a candidate word for substitution \newtext{(binned by quartiles)}, with 95\% asymptotic confidence intervals \newtext{(Goodman-based multinomial)}.
    \todo{Use log-y}
    }
    \label{fig:feature-susceptibilities-global}
\end{figure*}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/computed-figures/paper-variations_regression-globals_to_globalsvariation.png}
    \caption{\newtext{\textbf{Coefficients of the regressions of feature variations:}
	\todo{only aoa and clustering are impacted by frequency, the rest is only impacted by itself}
	}}
    \label{fig:feature-variations_regression-global}
\end{figure}

Second, the comparison with $\nu_{\phi}^0$ \newtext{and $\nu_{\phi}^{00}$} shows that there are two classes of attractors, depending on whether:
\begin{APAenumerate}
\item there is a triple intersection (of $y=x$, $\nu_{\phi}$, \newtext{and $\nu_{\phi}^0$ or $\nu_{\phi}^{00}$});
\item or $\nu_{\phi}$ always remains above or below $\nu_{\phi}^0$ \newtext{and $\nu_{\phi}^{00}$}.
\end{APAenumerate}
The first class (Number of synonyms and \newtext{Orthographic neighborhood density}) are features for which the substitution process only brings words slightly closer to $\nu_{\phi}^0$ \newtext{(for Number of synonyms) or $\nu_{\phi}^{00}$ (for Orthographic neighborhood density)}, and no uniform bias can be observed.
The second class (comprising Word frequency, Age of acquisition, Clustering coefficient\newtext{, and Number of letters}) are features for which the substitution process has a clear bias, positive or negative, with respect to \newtext{both} the purely random situation ($\mathcal{H}_0$) \newtext{and the semantically plausible random situation ($\mathcal{H}_{00}$)}.

Word frequency, with $\nu_{\phi}$ always significantly above $\nu_{\phi}^0$ \newtext{and $\nu_{\phi}^{00}$}, exhibits a strong bias towards more frequent words.
This, in turn, is consistent with the hypothesis that substitution is a recall process, since common words are favored over awkward ones.%, while it goes against the idea that it could be a familiarity process, where awkward terms would be favored.
Age of acquisition, Clustering coefficient \newtext{and Number of letters}, on the other hand, exhibit a clear negative bias for the substitution process (except for high clustering values or very high number of letters).
\newtext{The three} curves are significantly below their respective $\nu_{\phi}^0$ \newtext{and $\nu_{\phi}^{00}$ curves for most start values}, which is consistent with the literature on recall:
words learned earlier, with lower clustering coefficient \newtext{or with fewer letters} are easier to produce than average~\citep{nelson_how_2013,zevin_age_2002,baddeley_word_1975}.

To make things concrete, here is an example substitution taking place in the data set.
\newtext{Around mid-November 2008, several media websites reported the following quote from Burmese poet Saw Wai (arrested for one of his poems),
\begin{quote}
    ``Senior general Than Shwe is foolish with power.''
\end{quote}
and a smaller number of media websites, and blogs, reported the following,
\begin{quote}
    ``Senior general Than Shwe is \textbf{crazy} with power.''
\end{quote}
The word \emph{foolish} is acquired at an average of 8.94 years old, appears 675 times in the data set, has a Clustering coefficient of $8.2 \times 10^{-3}$ and is 7 letters long.
The word it was replaced with, \emph{crazy}, is acquired on average at 5.22 years old, appears about 4.1k times in the data set, has a Clustering coefficient of $1.7 \times 10^{-3}$, and is 5 letters long.}
Such a change, though minor in appearance, is a typical example of alteration along the lines shown by our results.

% foolish / crazy
%  frequency: (6.5147126908725301, 8.3192299386323256)
%  aoa: (8.94, 5.22)
%  clustering: (-4.7983132551456364, -6.369204119494885)
%  letters_count: (7, 5)

@@

\begin{figure*}[!th]
    \centering
    \includegraphics[width=\textwidth]{images/computed-figures/paper-variations-quantilebins_sentencerel.png}
    \caption{\todo{Bring back to 1, and use log-y}}
    \label{fig:feature-variations-sentencerel}
\end{figure*}
